---
title: "Data exploration"
output: html_document
date: "2024-08-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(data.table)
library(metR)
library(lubridate)
```


[Multi-Satellite Volcanic Sulfur Dioxide L4 Long-Term Global Database V4 (MSVOLSO2L4)](https://disc.gsfc.nasa.gov/datasets/MSVOLSO2L4_4/summary)

| Variable name | Description |
|-----------|------------------|
| volcano   | Name of volcano. |
| lat       | Latitude of volcano. |
| lon       | Longitude of volcano. |
| v_alt     | Altitude of volcano (km). |
| yyy       | Eruption year. |
| mm        | Eruption month of year. |
| dd        | Eruption day of month. |
| type      | Eruption style: exp = explosive, eff = effusive. |
| vei       | Eruption volcanic explosivity index (nd = no data or undetermined). |
| p_alto_obs| Observed plume altitude (km) where known. |
| p_alt_est | Estimated plume altitude (km) above vent: 10 km for explosive, 5 km for effusive. |
| so2_kt    | Measured SO2 mass in kilotons (= 1000 metric tons) |

```{r}
so2 <- readr::read_lines(here::here("data/MSVOLSO2L4_v04-00-2024m0129.txt"), skip = 48)  |>
  gsub(" +\\t", "\t", x = _) |>   # Líneas con espacios Y tab
  gsub(" +", "\t", x = _) |>      # Líneas con espacios
  gsub("\\t$", "", x = _) |>      # Tab extra al final
  fread(text = _, na.strings = c("nd", "-999")) |> 
  janitor::clean_names() |> 
  _[, date := lubridate::make_datetime(yyyy, mm, dd, tz = "UTC")]
```


Time series is not homogeneous because: 

* changes in sensors since Sep 2004: 
* better bandwidth (can separate better between O3 and SO2)
* smaller sensor footprint (more resolution)
* better algorithms? (validation with other satellite data)
* No data from TOMS between 1994/12/28 and 1996/07/15


```{r}
so2 |> 
  ggplot(aes(date, so2_kt)) +
  geom_point(aes(color = factor(vei))) +
  scale_color_viridis_d(option = "inferno", direction = -1) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  labs(x = "Date", y = "Measured SO2 mass in kilotons",
       color = "Explosivity index") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

SO2 that reaches the stratosphere produces a cooling effect in the troposphere (sulfuric acid aerosols are good at scattering the SW radiation == increase in the total albedo, also changes microphysic properties in deep convective clouds - smaller droplets). SO2 transport in the stratosphere depends on:
* location of the volcano (usually SO2 wont cross the Ecuator)
* Phase of the QBO -> influence meriodional wind shear, then the horizontal transport
* Altitude of the plume -> if it reach higher levels, it moves more quickly and more to the poles. 


If SO2 does not reaches the stratosphere it could produce produce a warming effect in the troposphere (why? chemical reactions?), it should be short as the aerosols are removed by rain.


weak eruptions have more *local* impacts? (don't know if we can see local effects, separate the effect of the eruptions from the regional variability)

```{r}
so2 |> 
  ggplot(aes(factor(vei), p_alt_obs)) +
  geom_point(position = "jitter", alpha = 0.5, color = "cyan4") +
  facet_wrap(~type) +
  labs(x = "Explocivity index", y = "Observed plume altitude (km)") +
  theme_minimal()
```

```{r}
map <- rnaturalearth::ne_coastline(scale = 10, returnclass = "sf")

so2 |> 
  ggplot(aes(lon, lat)) +
  geom_point(aes(color = p_alt_obs  >= 12)) +
  geom_sf(data = map, inherit.aes = FALSE, fill = NA) +
  labs(x =  NULL, y = NULL) +
  theme_minimal()

```

```{r}
library(patchwork)

so2 |> 
  _[, .(mean_so2 = sum(so2_kt)), by = .(yyyymm = lubridate::make_date(yyyy, mm))] |> 
  ggplot(aes(yyyymm, mean_so2)) +
  geom_line() +
  geom_hline(yintercept = 5000, linetype = 3) +
  labs(x = NULL, y = "SO2 (Kt)",
       color = "Explosivity index") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  
  so2 |> 
  _[, .(so2_total = sum(so2_kt),
        p_alt_obs = max(p_alt_obs, na.rm = TRUE)), by = .(yyyy, volcano, lat)] |> 
  ggplot(aes(yyyy, lat)) +
  geom_point(aes(size = p_alt_obs, fill = so2_total), shape = 21, alpha = 0.7) +
  scale_fill_viridis_c(option = "inferno", direction = -1,
                       guide = guide_colorbar(barwidth = 10,
                                              barheight = 0.5)) +
  scale_size(range = c(0.1, 8)) +
  labs(x = NULL, y = "Latitude", 
       size = "Observed plume\naltitude (km)", fill = "Total SO2 (Kt)") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  
  plot_layout(ncol = 1, heights = c(0.25, 0.75))
```


## ERA5

```{r}
file_list  <- Sys.glob("/g/data/rt52/era5/single-levels/monthly-averaged/2t/*/*")

#GlanceNetCDF(file_list[1])

# t2_mean <- purrr::map(file_list, function(l) {

#   t2 <- ReadNetCDF(l)
# print(l)
#   global_mean <- t2[, .(t2m = mean(t2m, na.rm = TRUE)), by = time]

#   austrlia_mean  <- t2[longitude %between% c(110, 155) & latitude %between% c(-45, 11), 
#   .(t2m = mean(t2m, na.rm = TRUE)), by = time]

#   list(global_mean = global_mean, austrlia_mean = austrlia_mean)
# })  |> 
#   purrr::reduce(function(x, y) list(global_mean = rbind(x$global_mean, y$global_mean),
#                              austrlia_mean = rbind(x$austrlia_mean, y$austrlia_mean)))


# readr::write_rds(t2_mean, "data/t2_mean.rds")

t2_mean <- readr::read_rds("data/t2_mean.rds")

global_mean  <- t2_mean$global_mean |>
  _[, t2m_a := t2m - mean(t2m), by = .(month(time))] |> 
  _[, time := as_datetime(time, tz = "UTC")]


```


```{r}
# group_eruptions <- function(time, threshold = 5) {
#   dt <- c(Inf, diff(time))
#   dt <- dplyr::if_else(dt > threshold, lubridate::as_date(time), NA)
#   data.table::nafill(dt, type = "locf")
# }

# calculate_tbase <- function(temperature_serie, ini_dates, period = 12) {
#   
#   purrr::map_dbl(ini_dates, function(ini_date) {
#     
#     start_period <- ini_date - months(period)
#     
#     as.numeric(temperature_serie[time %between% c(start_period, ini_date), 
#                                  .(base_t = mean(t2m_a))])
#   }) 
#   
# }
# 
# get_series <- function(ini_date) {
#   
#   start_time <- ini_date - years(1)
#   end_time <- ini_date + years(3)
#   temp <- global_mean[time %between% c(start_time, end_time)] 
#   
#   list(time = temp$time,
#        t2m_a = temp$t2m_a)
#   
# }
```


```{r}
eruptions <- so2[, ini_date := group_eruptions(date, 60), by = volcano] |> 
  _[, p_alt_obs := fifelse(is.na(p_alt_obs), p_alt_est ,p_alt_obs)] |> 
  _[, .(p_alt_obs = max(p_alt_obs, na.rm = TRUE),
        so2_kt = sum(so2_kt, na.rm = TRUE),
        lat = unique(lat),
        lon = unique(lon),
        vei = max(vei, na.rm = TRUE)), by = .(volcano, ini_date)] |>
  _[, let(base_t = calculate_tbase(global_mean, ini_date)),
    by = .(volcano, ini_date)] |>
  _[, let(id = paste0(volcano, "_", ini_date))] |> 
  _[]

temp_effect <-
  eruptions[,  get_series(ini_date),
            by = .(volcano, ini_date)] |>
  eruptions[i = _, on = c("volcano", "ini_date")] |> 
  _[, let(id = paste0(volcano, "_", ini_date),
          lag =  as.numeric(difftime(time, ini_date))/2592000)]
```


```{r}
eruptions |>
  ggplot(aes(ini_date, volcano)) +
  geom_point(aes(color = p_alt_obs > 20)) 
```

Neither so2 > 5000 or p_alt_obs > 20 km is good enough to clasify eruptions as "strong". Or at least not all eruptions that fall into this categories show a cooling effect on surface temperature anomaly (global). 

So, can we use both? Or maybe look for the eruptios that shows this cooling effect. 
```{r}
temp_effect |>
  _[so2_kt > 5000] |>
  # _[, .N, by = id] |>
  ggplot(aes(lag, t2m_a - base_t)) +
  geom_vline(xintercept = 0, color = "darkorange") +
  geom_hline(yintercept = 0, color = "grey") +
  geom_line(aes(group = id), alpha = 1) +
  geom_area(fill = "grey", alpha = 0.5) +
  facet_wrap(~volcano, ncol = 1) +
  labs(x = "Months since eruption",
       y = "T 2m anomaly") +
  theme_minimal()

temp_effect |>
  _[p_alt_obs > 20] |>
  ggplot(aes(lag, t2m_a - base_t)) +
  geom_vline(xintercept = 0, color = "darkorange") +
    geom_hline(yintercept = 0,color = "grey") +
  geom_line(aes(group = id)) +
  geom_smooth(method = "lm", alpha = 0.5, linewidth = 0.5) +

  facet_wrap(~volcano, ncol = 3) +
  labs(x = "Months since eruption",
       y = "T 2m anomaly") +
  theme_minimal()

```

I can calculate the slop for the trend associated with each time series (for each eruption). 

```{r}
temp_slop <- temp_effect |>
  _[, FitLm(t2m_a - base_t, lag), by = id] |>
  _[term == "lag"] |>
  _[eruptions, on = .NATURAL]

strong <- temp_slop[estimate < -0.006, .(id, volcano, estimate, so2_kt, p_alt_obs, vei)]
```

There are `r temp_slop[estimate < 0, .N, ]` eruptions with a negative slop. Although many of them have a very small slop.

```{r}
temp_slop |>
  ggplot(aes(estimate)) +
  geom_histogram(binwidth = 0.001) +
  geom_vline(xintercept = 0, color = "darkorange") +
  geom_vline(xintercept = -0.006, color = "darkorange") +
  annotate("text", y = 10, x = -0.012, label = "Strong\n eruptions?") +
  labs(x = "Slop", y = NULL) +
  theme_minimal()
```



```{r}
temp_slop |>
  ggplot(aes(estimate, so2_kt)) +
  geom_vline(xintercept = 0, color = "darkorange") +
  geom_point(aes(color = p_alt_obs)) +
  # geom_smooth(method = "lm", alpha = 0.5) +
  geom_label(data = strong, aes(label = volcano),
             nudge_y = 0.2) +
  scale_color_viridis_c(direction = -1) +
  scale_y_log10() +
  labs(x = "Slope", y = "SO2 emission", color = "Plume altitud") +
  theme_minimal()
```


```{r}
temp_effect[id %in% strong$id] |> 
  _[] |> 
  ggplot(aes(lag, t2m_a - base_t)) +
  geom_vline(xintercept = 0, color = "darkorange") +
    geom_hline(yintercept = 0,color = "grey") +
  geom_line(aes(group = id)) +
  geom_smooth(method = "lm", alpha = 0.5, linewidth = 0.5) +
  geom_text(data = strong, x = -10, y = -0.5, hjust = 0, size = 3,
            aes(label = paste0("SO2: ", so2_kt, "\nAlt: ", p_alt_obs, "\nVEI: ", vei))) +
  facet_wrap(~id, ncol = 3) +
  labs(x = "Months since eruption",
       y = "T 2m anomaly") +
  theme_minimal()
```

