---
title: "Data exploration"
output: html_document
date: "2024-08-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(data.table)
library(metR)
library(lubridate)
```


[Multi-Satellite Volcanic Sulfur Dioxide L4 Long-Term Global Database V4 (MSVOLSO2L4)](https://disc.gsfc.nasa.gov/datasets/MSVOLSO2L4_4/summary)

| Variable name | Description |
|-----------|------------------|
| volcano   | Name of volcano. |
| lat       | Latitude of volcano. |
| lon       | Longitude of volcano. |
| v_alt     | Altitude of volcano (km). |
| yyy       | Eruption year. |
| mm        | Eruption month of year. |
| dd        | Eruption day of month. |
| type      | Eruption style: exp = explosive, eff = effusive. |
| vei       | Eruption volcanic explosivity index (nd = no data or undetermined). |
| p_alto_obs| Observed plume altitude (km) where known. |
| p_alt_est | Estimated plume altitude (km) above vent: 10 km for explosive, 5 km for effusive. |
| so2_kt    | Measured SO2 mass in kilotons (= 1000 metric tons) |

```{r}
so2 <- readr::read_lines(here::here("data/MSVOLSO2L4_v04-00-2024m0129.txt"), skip = 48)  |>
  gsub(" +\\t", "\t", x = _) |>   # Líneas con espacios Y tab
  gsub(" +", "\t", x = _) |>      # Líneas con espacios
  gsub("\\t$", "", x = _) |>      # Tab extra al final
  fread(text = _, na.strings = c("nd", "-999")) |> 
  janitor::clean_names() |> 
  _[, date := lubridate::make_datetime(yyyy, mm, dd, tz = "UTC")]
```


Time series is not homogeneous because: 

* changes in sensors since Sep 2004: 
* better bandwidth (can separate better between O3 and SO2)
* smaller sensor footprint (more resolution)
* better algorithms? (validation with other satellite data)
* No data from TOMS between 1994/12/28 and 1996/07/15


```{r}
so2 |> 
  ggplot(aes(date, so2_kt)) +
  geom_point(aes(color = factor(vei))) +
  scale_color_viridis_d(option = "inferno", direction = -1) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  labs(x = "Date", y = "Measured SO2 mass in kilotons",
       color = "Explosivity index") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

SO2 that reaches the stratosphere produces a cooling effect in the troposphere (sulfuric acid aerosols are good at scattering the SW radiation == increase in the total albedo, also changes microphysic properties in deep convective clouds - smaller droplets). SO2 transport in the stratosphere depends on:
* location of the volcano (usually SO2 wont cross the Ecuator)
* Phase of the QBO -> influence meriodional wind shear, then the horizontal transport
* Altitude of the plume -> if it reach higher levels, it moves more quickly and more to the poles. 


If SO2 does not reaches the stratosphere it could produce produce a warming effect in the troposphere (why? chemical reactions?), it should be short as the aerosols are removed by rain.


weak eruptions have more *local* impacts? (don't know if we can see local effects, separate the effect of the eruptions from the regional variability)

```{r}
so2 |> 
  ggplot(aes(factor(vei), p_alt_obs)) +
  geom_point(position = "jitter", alpha = 0.5, color = "cyan4") +
  facet_wrap(~type) +
  labs(x = "Explocivity index", y = "Observed plume altitude (km)") +
  theme_minimal()
```

```{r}
map <- rnaturalearth::ne_coastline(scale = 10, returnclass = "sf")

so2 |> 
  ggplot(aes(lon, lat)) +
  geom_point(aes(color = p_alt_obs  >= 12)) +
  geom_sf(data = map, inherit.aes = FALSE, fill = NA) +
  labs(x =  NULL, y = NULL) +
  theme_minimal()

```

```{r}
library(patchwork)

so2 |> 
  _[, .(mean_so2 = sum(so2_kt)), by = .(yyyymm = lubridate::make_date(yyyy, mm))] |> 
  ggplot(aes(yyyymm, mean_so2)) +
  geom_line() +
  geom_hline(yintercept = 5000, linetype = 3) +
  labs(x = NULL, y = "SO2 (Kt)",
       color = "Explosivity index") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  
  so2 |> 
  _[, .(so2_total = sum(so2_kt),
        p_alt_obs = max(p_alt_obs, na.rm = TRUE)), by = .(yyyy, volcano, lat)] |> 
  ggplot(aes(yyyy, lat)) +
  geom_point(aes(size = p_alt_obs, fill = so2_total), shape = 21, alpha = 0.7) +
  scale_fill_viridis_c(option = "inferno", direction = -1,
                       guide = guide_colorbar(barwidth = 10,
                                              barheight = 0.5)) +
  scale_size(range = c(0.1, 8)) +
  labs(x = NULL, y = "Latitude", 
       size = "Observed plume\naltitude (km)", fill = "Total SO2 (Kt)") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  
  plot_layout(ncol = 1, heights = c(0.25, 0.75))
```


## ERA5

```{r}
file_list  <- Sys.glob("/g/data/rt52/era5/single-levels/monthly-averaged/2t/*/*")

#GlanceNetCDF(file_list[1])

# t2_mean <- purrr::map(file_list, function(l) {

#   t2 <- ReadNetCDF(l)
# print(l)
#   global_mean <- t2[, .(t2m = mean(t2m, na.rm = TRUE)), by = time]

#   austrlia_mean  <- t2[longitude %between% c(110, 155) & latitude %between% c(-45, 11), 
#   .(t2m = mean(t2m, na.rm = TRUE)), by = time]

#   list(global_mean = global_mean, austrlia_mean = austrlia_mean)
# })  |> 
#   purrr::reduce(function(x, y) list(global_mean = rbind(x$global_mean, y$global_mean),
#                              austrlia_mean = rbind(x$austrlia_mean, y$austrlia_mean)))


# readr::write_rds(t2_mean, "data/t2_mean.rds")

t2_mean <- readr::read_rds("data/t2_mean.rds")

global_mean  <- t2_mean$global_mean |>
  _[, t2m_a := t2m - mean(t2m), by = .(month(time))] |> 
  _[, time := as_datetime(time, tz = "UTC")]


```


```{r}
eruptions <- so2[, ini_date := group_eruptions(date, 60), by = volcano] |> 
  _[, p_alt_obs := fifelse(is.na(p_alt_obs), p_alt_est ,p_alt_obs)] |> 
  _[, .(p_alt_obs = max(p_alt_obs, na.rm = TRUE),
        so2_kt = sum(so2_kt, na.rm = TRUE),
        lat = unique(lat),
        lon = unique(lon),
        vei = max(vei, na.rm = TRUE)), by = .(volcano, ini_date)] |>
  # _[, let(base_t = calculate_tbase(global_mean, ini_date)),
  #   by = .(volcano, ini_date)] |>
  _[, let(id = paste0(volcano, "_", ini_date))] |> 
  _[]

temp_effect <-
  eruptions[,  get_series(ini_date),
            by = .(volcano, ini_date)] |>
  eruptions[i = _, on = c("volcano", "ini_date")] |> 
  _[, let(id = paste0(volcano, "_", ini_date),
          lag =  as.numeric(difftime(time, ini_date))/2592000)]
```


Neither so2 > 5000 or p_alt_obs > 20 km is good enough to clasify eruptions as "strong". Or at least not all eruptions that fall into this categories show a cooling effect on surface temperature anomaly (global). 

So, can we use both? Or maybe look for the eruptios that shows this cooling effect. 
```{r}
temp_effect |>
  _[so2_kt > 5000] |>
  # _[, .N, by = id] |>
  ggplot(aes(lag, t2m_a - base_t)) +
  geom_vline(xintercept = 0, color = "darkorange") +
  geom_hline(yintercept = 0, color = "grey") +
  geom_line(aes(group = id), alpha = 1) +
  geom_area(fill = "grey", alpha = 0.5) +
  facet_wrap(~volcano, ncol = 1) +
  labs(x = "Months since eruption",
       y = "T 2m anomaly") +
  theme_minimal()

temp_effect |>
  _[p_alt_obs > 20] |>
  ggplot(aes(lag, t2m_a - base_t)) +
  geom_vline(xintercept = 0, color = "darkorange") +
    geom_hline(yintercept = 0,color = "grey") +
  geom_line(aes(group = id)) +
  geom_smooth(method = "lm", alpha = 0.5, linewidth = 0.5) +

  facet_wrap(~volcano, ncol = 3) +
  labs(x = "Months since eruption",
       y = "T 2m anomaly") +
  theme_minimal()

```

I can calculate the slop for the trend associated with each time series (for each eruption). 

```{r}
temp_slop <- temp_effect |>
  _[, FitLm(t2m_a - base_t, lag), by = id] |>
  _[term == "lag"] |>
  _[eruptions, on = .NATURAL]

strong <- temp_slop[estimate < -0.006, .(id, volcano, estimate, so2_kt, p_alt_obs, vei)]
```

There are `r temp_slop[estimate < 0, .N, ]` eruptions with a negative slop. Although many of them have a very small slop.

```{r}
temp_slop |>
  ggplot(aes(estimate)) +
  geom_histogram(binwidth = 0.001) +
  geom_vline(xintercept = 0, color = "darkorange") +
  geom_vline(xintercept = -0.006, color = "darkorange") +
  annotate("text", y = 10, x = -0.012, label = "Strong\n eruptions?") +
  labs(x = "Slop", y = NULL) +
  theme_minimal()
```



```{r}
temp_slop |>
  ggplot(aes(estimate, so2_kt)) +
  geom_vline(xintercept = 0, color = "darkorange") +
  geom_point(aes(color = p_alt_obs)) +
  # geom_smooth(method = "lm", alpha = 0.5) +
  geom_label(data = strong, aes(label = volcano),
             nudge_y = 0.2) +
  scale_color_viridis_c(direction = -1) +
  scale_y_log10() +
  labs(x = "Slope", y = "SO2 emission", color = "Plume altitud") +
  theme_minimal()
```

Nice, it didn't work. Almost all the eruptions I got are actually not that strong, my strategy doesn't account for mild eruptions close to a strong one. In my case I getting all the small eruptions in 1991 around Pinatubo. 

Comming back to the bases. What are other people doing?

```{r}
temp_effect[id %in% strong$id] |> 
  _[] |> 
  ggplot(aes(lag, t2m_a - base_t)) +
  geom_vline(xintercept = 0, color = "darkorange") +
    geom_hline(yintercept = 0,color = "grey") +
  geom_line(aes(group = id)) +
  geom_smooth(method = "lm", alpha = 0.5, linewidth = 0.5) +
  geom_text(data = strong, x = -10, y = -0.5, hjust = 0, size = 3,
            aes(label = paste0("SO2: ", so2_kt, "\nAlt: ", p_alt_obs, "\nVEI: ", vei))) +
  facet_wrap(~id, ncol = 3) +
  labs(x = "Months since eruption",
       y = "T 2m anomaly") +
  theme_minimal()
```

Let's use a *multi-factor* criteria including the SO2 emission (over 900 kt), the altitude of the plume (over 15 km) and the  Eruption volcanic explosivity index (less important as the other conditions makes this index high enough). 

```{r}
# eruptions |>
#     _[vei > 3] |> 
#   ggplot(aes(ini_date, volcano)) +
#   geom_point(aes(fill = p_alt_obs > 18, color = so2_kt > 3000, size = so2_kt), shape = 21) +
#   scale_color_manual(values = c("white", "black")) +
#   scale_fill_manual(values = c("grey", "darkorange")) +
#   theme_minimal()

eruptions |> 
  _[so2_kt > 900] |> 
  ggplot(aes(so2_kt)) +
  geom_histogram(binwidth = 100) +
  geom_label(data = ~.x[, .(.N, so2 = max(so2_kt))], y = 2, aes(label = N, x = so2)) +
  labs(x = "Emissions of SO2") +
  theme_minimal() +

eruptions |> 
  _[p_alt_obs > 15] |> 
  ggplot(aes(p_alt_obs)) +
  geom_histogram(binwidth = 0.5) +
  geom_label(data = ~.x[, .(.N, alt = max(p_alt_obs))], y = 18, aes(label = N, x = alt)) +
  labs(x = "Max altitude of plume") +
  theme_minimal()

new_strong <- eruptions |> 
  _[so2_kt > 900 & p_alt_obs > 15 & vei >= 5] |> 
  _[]

new_strong |> 
  readr::write_rds("data/strong_volcanos.rds")
```

We get 7 eruptions that were high enough to reach the stratosphere and/or emitted enough SO2 to affect the temperature (localy or globaly). 

Checking if we can see and effect on the global temperature anomaly. If not, does not mean that the eruption didn't have an effect, we could have regional effect that cancel out. 

```{r}
temp_effect[id %in% new_strong$id] |> 
  _[] |> 
  ggplot(aes(lag, t2m_a - base_t)) +
  geom_vline(xintercept = 0, color = "darkorange") +
    geom_hline(yintercept = 0,color = "grey") +
  geom_line(aes(group = id)) +
  geom_smooth(method = "lm", alpha = 0.5, linewidth = 0.5) +
  geom_text(data = new_strong, x = 0, y = -0.4, hjust = 0, size = 3,
            aes(label = paste0("SO2: ", so2_kt, "\nAlt: ", p_alt_obs, "\nVEI: ", vei))) +
  facet_wrap(~id, ncol = 3) +
  labs(x = "Months since eruption",
       y = "T 2m anomaly") +
  theme_minimal()
```
Only El Chichon, Mt Pinatubo and Cerro Hudson show a negative slop in the temperature anomaly comparing with the mean temperature for the year before the eruption. The Cerro Hudson signal may be mixed with Pinatubo. 

Choosing  a different base period can also affect this results. hansen1996 mentions that a 2 or a 5 years period could also be used.

But... it looks like the same. 


```{r}
base_period = 2 # years

eruptions_5 <- so2[, ini_date := group_eruptions(date, 60), by = volcano] |> 
  _[, p_alt_obs := fifelse(is.na(p_alt_obs), p_alt_est ,p_alt_obs)] |> 
  _[, .(p_alt_obs = max(p_alt_obs, na.rm = TRUE),
        so2_kt = sum(so2_kt, na.rm = TRUE),
        lat = unique(lat),
        lon = unique(lon),
        vei = max(vei, na.rm = TRUE)), by = .(volcano, ini_date)] |>
  _[, let(base_t = calculate_tbase(global_mean, ini_date, period = base_period*12)),
    by = .(volcano, ini_date)] |>
  _[, let(id = paste0(volcano, "_", ini_date))] |> 
  _[]

temp_effect_5 <-
  eruptions_5[,  get_series(ini_date, before = base_period),
            by = .(volcano, ini_date)] |>
  eruptions_5[i = _, on = c("volcano", "ini_date")] |> 
  _[, let(id = paste0(volcano, "_", ini_date),
          lag =  as.numeric(difftime(time, ini_date))/2592000)]


temp_effect_5[id %in% new_strong$id] |> 
  _[] |> 
  ggplot(aes(lag, t2m_a - base_t)) +
  geom_vline(xintercept = 0, color = "darkorange") +
    geom_hline(yintercept = 0,color = "grey") +
  geom_line(aes(group = id)) +
  geom_smooth(method = "lm", alpha = 0.5, linewidth = 0.5) +
  geom_text(data = new_strong, x = 0, y = -0.4, hjust = 0, size = 3,
            aes(label = paste0("SO2: ", so2_kt, "\nAlt: ", p_alt_obs, "\nVEI: ", vei))) +
  facet_wrap(~id, ncol = 3) +
  labs(x = "Months since eruption",
       y = "T 2m anomaly") +
  theme_minimal()
```

