---
title: "ERA5 - extremes"
output: html_document
date: "2024-08-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(data.table)
library(metR)
library(lubridate)
library(unglue)

source("scripts/help_functions.R")
```

## Strong volcanos

```{r}
so2 <- readr::read_lines(here::here("data/MSVOLSO2L4_v04-00-2024m0129.txt"), skip = 48)  |>
  gsub(" +\\t", "\t", x = _) |>   # Líneas con espacios Y tab
  gsub(" +", "\t", x = _) |>      # Líneas con espacios
  gsub("\\t$", "", x = _) |>      # Tab extra al final
  fread(text = _, na.strings = c("nd", "-999")) |> 
  janitor::clean_names() |> 
  _[, date := lubridate::make_datetime(yyyy, mm, dd, tz = "UTC")]

strong <- so2[, ini_date := group_eruptions(date, 60), by = volcano] |> 
  _[, p_alt_obs := fifelse(is.na(p_alt_obs), p_alt_est ,p_alt_obs)] |> 
  _[, .(p_alt_obs = max(p_alt_obs, na.rm = TRUE),
        so2_kt = sum(so2_kt, na.rm = TRUE),
        lat = unique(lat),
        lon = unique(lon),
        vei = max(vei, na.rm = TRUE)), by = .(volcano, ini_date)] |>
  # _[, let(base_t = calculate_tbase(global_mean, ini_date)),
  #   by = .(volcano, ini_date)] |>
  _[, let(id = paste0(volcano, "_", ini_date))] |> 
  _[(so2_kt > 900 & p_alt_obs > 15 & vei >= 5) | stringr::str_detect(volcano, "Hunga_Tonga") & p_alt_obs > 15] |> 
  _[]

map <- rnaturalearth::ne_coastline(scale = 50, returnclass = "sf")

strong |> 
  ggplot(aes(lon, lat)) +
  geom_sf(data = map, inherit.aes = FALSE, fill = NA, color = "grey") +
  geom_point(aes(fill = p_alt_obs, size = so2_kt), shape = 21) +
  geom_label(aes(label = volcano), nudge_y = 12) +
  scale_fill_viridis_c(guide = guide_colorbar(barwidth = 10,
                                              barheight = 0.5)) +
  labs(x =  NULL, y = NULL, fill = "Altitude", size = "SO2") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
## Temperature anomaly after volcanic eruptions

Global temperature time series without global warming trend and annual cycle. 

```{r}
enso <- rsoi::download_oni(use_cache = TRUE) |> 
  as.data.table() |> 
  _[,.(time = as.POSIXct(Date), oni = dSST3.4)] 
```


```{r}
t2_mean <- readr::read_rds("data/t2_mean_daily.rds") |> 
  _[, time := floor_date(time, "day")] |> 
  _[, t2m_notrend := pracma::detrend(t2m)] |> 
  _[, t2m_a := t2m_notrend - mean(t2m_notrend), by = .(day(time), month(time))]
# |>
#   enso[i = _, on = "time"] |> 
#   _[, oni := nafill(oni, "locf")] |> 
#   _[, t2m_a := resid(lm(t2m_noannual ~ oni))]


t2_mean |> 
  ggplot(aes(time, t2m_noannual)) +
  geom_line() +
  geom_line(aes(y = t2m_a), color = "orange") +
  labs(x = "Years", y = "2m temperature anomaly [C]") +
  theme_minimal()
```

Now, define the eruption and no-eruption period to calculate a base temperature using the complete time series. An eruption period will include the 24 months after one of the selected eruptions. 

```{r}
t2_mean[, period := fcase(time %between% c(strong$ini_date[1], strong$ini_date[1] + months(24)), "eruption",
                          time %between% c(strong$ini_date[2], strong$ini_date[2] + months(24)), "eruption",
                          time %between% c(strong$ini_date[3], strong$ini_date[3] + months(24)), "eruption",
                          time %between% c(strong$ini_date[4], strong$ini_date[4] + months(24)), "eruption",
                          time %between% c(strong$ini_date[5], strong$ini_date[5] + months(24)), "eruption",
                          default = "no eruption")]

t2_mean |> 
  _[, time := as_date(time, tz = "GMT")] |>
  _[] |> 
  ggplot(aes(time, t2m_a)) +
  geom_line(aes(color = period)) +
  geom_point(data = strong, aes(x = ini_date), y = 0) +
  geom_hline(data = t2_mean[, mean(t2m_a), by = period], aes(yintercept = V1, color = period)) +
  labs(x = "Years", y = "2m temperature anomaly [C]") +
  theme_minimal()
```


```{r}
t2_mean[, mean(t2m_a), by = period]

t2_mean[time %between% as_date(c("1976-01-01", "1981-12-31")), mean(t2m_a)]

t2m_noeruption_base <- strong[, get_series(t2_mean, ini_date), by = .(volcano, ini_date)] |> 
  strong[i = _, on = c("volcano", "ini_date")] |> 
  _[, let(id = paste0(volcano, "_", ini_date),
          lag =  as.integer(difftime(time, ini_date)/86400),
          diff = t2m_a - 0.04829714)] |> 
  _[]

t2m_noeruption_base |> 
   _[, .(t_mean = mean(diff)), by = .(lag, id)] |> 
  _[, let(t_mean = frollmean(t_mean, 30, align = "center"))] |> 
  _[lag %between% c(0, 365*1.5)] |> 
  ggplot(aes(lag, t_mean)) +
  geom_hline(yintercept = 0) +
  geom_line(aes(color = id), alpha = 0.5) +
  # geom_line(data = t2m_noeruption_base[lag %between% c(0, 365*1.5), .(t_mean = mean(diff)), by = lag], aes(y = t_mean)) +
  geom_line(data = t2m_noeruption_base[lag %between% c(-15, 365*1.5+15), .(t_mean = mean(diff)), by = lag] |> 
              _[, let(t_mean = frollmean(t_mean, 30, align = "center"))], aes(y = t_mean)) +
  # geom_smooth() +
  # geom_smooth(aes(color = id))
  guides(color = guide_legend(nrow = 2)) +
  # facet_wrap(~id) +
  scale_x_continuous(breaks = seq(0, 550, 60)) +
  labs(x = "Days since eruption", y = "t2m anomaly - base",
       color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.box="vertical")
```
```{r}
tropics_non_ve <- readr::read_rds("data/t2_mean_daily_regions.rds")$tropics |> 
  _[, time := floor_date(time, "day")] |> 
  _[, t2m_notrend := pracma::detrend(t2m)] |> 
  _[, t2m_a := t2m_notrend - mean(t2m_notrend), by = .(day(time), month(time))]

tropics_non_ve[, period := fcase(time %between% c(strong$ini_date[1], strong$ini_date[1] + months(24)), "eruption",
                          time %between% c(strong$ini_date[2], strong$ini_date[2] + months(24)), "eruption",
                          time %between% c(strong$ini_date[3], strong$ini_date[3] + months(24)), "eruption",
                          time %between% c(strong$ini_date[4], strong$ini_date[4] + months(24)), "eruption",
                          time %between% c(strong$ini_date[5], strong$ini_date[5] + months(24)), "eruption",
                          default = "no eruption")]

tropics_non_ve[, mean(t2m_a), by = period]

tropics_noeruption_base <- strong[, get_series(tropics_non_ve, ini_date), by = .(volcano, ini_date)] |> 
  strong[i = _, on = c("volcano", "ini_date")] |> 
  _[, let(id = paste0(volcano, "_", ini_date),
          lag =  as.integer(difftime(time, ini_date)/86400),
          diff = t2m_a - -0.001320377	)] |> 
  _[]

tropics_noeruption_base |> 
   _[, .(t_mean = mean(diff)), by = .(lag, id)] |> 
  _[, let(t_mean = frollmean(t_mean, 30, align = "center"))] |> 
  _[lag %between% c(0, 365*1.5)] |> 
  ggplot(aes(lag, t_mean)) +
  geom_hline(yintercept = 0) +
  geom_line(aes(color = id), alpha = 0.5) +
  # geom_line(data = t2m_noeruption_base[lag %between% c(0, 365*1.5), .(t_mean = mean(diff)), by = lag], aes(y = t_mean)) +
  geom_line(data = tropics_noeruption_base[lag %between% c(-15, 365*1.5+15), .(t_mean = mean(diff)), by = lag] |> 
              _[, let(t_mean = frollmean(t_mean, 30, align = "center"))], aes(y = t_mean)) +
  # geom_smooth() +
  # geom_smooth(aes(color = id))
  guides(color = guide_legend(nrow = 2)) +
  # facet_wrap(~id) +
  scale_x_continuous(breaks = seq(0, 550, 60)) +
  labs(x = "Days since eruption", y = "t2m anomaly - base",
       color = NULL, title = "Tropics") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.box="vertical")
```

Using 12 months previous to the eruption as a base period. 

```{r}

t2m_2y <- strong[, let(base = calculate_tbase(t2_mean, ini_date, period = 12)),
                 by = .(volcano, ini_date)] |>
  _[, get_series(t2_mean, ini_date, before = 1, after = 2), by = .(volcano, ini_date)] |> 
  strong[i = _, on = c("volcano", "ini_date")] |> 
  _[, let(id = paste0(volcano, "_", ini_date),
          lag =  as.numeric(difftime(time, ini_date)/86400),
          diff = t2m_a - base)] |> 
  _[]

t2m_2y |> 
  _[lag %between% c(0, 365*1.5)] |> 
  ggplot(aes(lag, diff)) +
  geom_hline(yintercept = 0) +
  geom_line(aes(color = id), alpha = 0.5) +
  # geom_line(data = t2m_2y[lag %between% c(0, 365*1.5), .(t_mean = mean(diff)), by = lag], aes(y = t_mean)) +
  geom_line(data = t2m_2y[lag %between% c(-15, 365*1.5+15), .(t_mean = mean(diff)), by = lag] |> 
              _[, let(t_mean = frollmean(t_mean, 30, align = "center"))], aes(y = t_mean)) +
  # geom_smooth() +
  # geom_smooth(aes(color = id))
  guides(color = guide_legend(nrow = 2)) +
  facet_wrap(~id) +
  scale_x_continuous(breaks = seq(0, 550, 60)) +
  labs(x = "Days since eruption", y = "t2m anomaly - base",
       color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.box="vertical")

t2m_2y |> 
  _[, .(t_mean = mean(diff)), by = .(lag, id)] |> 
  _[, let(t_mean = frollmean(t_mean, 30, align = "center"))] |> 
  _[lag %between% c(0, 365*1.5)] |> 
  ggplot(aes(lag, t_mean)) +
  geom_hline(yintercept = 0, color =  "grey") +
  annotate("rect", xmin = 120, xmax = 240, ymin = -Inf, ymax = Inf, alpha = .1) +
  geom_line(aes(color = id), alpha = 0.5) +
  geom_line(data = t2m_2y[lag %between% c(-15, 365*1.5+15) & volcano != "Hunga_Tonga_Hungaapai", 
                          .(t_mean = mean(diff)), by = lag] |> 
              _[, let(t_mean = frollmean(t_mean, 30, align = "center"))], aes(y = t_mean)) +
  # geom_smooth() +
  guides(color = guide_legend(nrow = 2)) +
  # facet_wrap(~id) +
  scale_x_continuous(breaks = seq(0, 550, 60)) +
  labs(x = "Days since eruption", y = "t2m anomaly - base",
       color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.box="vertical")

t2m_2y[lag %between% c(0, 365*1), .(t_mean = mean(diff)), by = lag] |> 
  ggplot(aes(lag, t_mean)) +
  geom_line() +
  geom_line(data = t2m_noeruption_base[lag %between% c(0, 365*1), .(t_mean = mean(diff)), by = lag], color = "orange")
```

```{r}
tropics <- readr::read_rds("data/t2_mean_daily_regions.rds")$tropics |> 
  _[, time := floor_date(time, "day")] |> 
  _[, t2m_notrend := pracma::detrend(t2m)] |> 
  _[, t2m_a := t2m_notrend - mean(t2m_notrend), by = .(day(time), month(time))]


tropics <- strong[, let(base = calculate_tbase(tropics, ini_date, period = 12)),
                 by = .(volcano, ini_date)] |>
  _[, get_series(tropics, ini_date, before = 1, after = 2), by = .(volcano, ini_date)] |> 
  strong[i = _, on = c("volcano", "ini_date")] |> 
  _[, let(id = paste0(volcano, "_", ini_date),
          lag =  as.numeric(difftime(time, ini_date)/86400),
          diff = t2m_a - base)] |> 
  _[]

tropics |> 
   _[, .(t_mean = mean(diff)), by = .(lag, id)] |> 
  _[, let(t_mean = frollmean(t_mean, 30, align = "center"))] |> 
  _[lag %between% c(0, 365*1.5)] |> 
  ggplot(aes(lag, t_mean)) +
  geom_hline(yintercept = 0, color =  "grey") +
  annotate("rect", xmin = 120, xmax = 240, ymin = -Inf, ymax = Inf, alpha = .1) +
  geom_line(aes(color = id), alpha = 0.5) +
  geom_line(data = tropics[lag %between% c(-15, 365*1.5+15), 
                          .(t_mean = mean(diff)), by = lag] |> 
              _[, let(t_mean = frollmean(t_mean, 30, align = "center"))], aes(y = t_mean)) +
  # geom_smooth() +
  guides(color = guide_legend(nrow = 2)) +
  # facet_wrap(~id) +
  scale_x_continuous(breaks = seq(0, 550, 60)) +
  labs(x = "Days since eruption", y = "t2m anomaly - base",
       color = NULL, title = "Extratropics SH") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.box="vertical")


```



### Impact in T2m fields

Reads 30 days around day 180 after each eruption and calculate the mean. I'm using the daily temperature anomaly (no long term trend or seasonal cycle).

```{r}
times <- (strong$ini_date + days(121)) |> 
  purrr::map(function(d) {
    
    # c(d - days(1), d + days(1))
    times <- list(seq.Date(d - days(15), d + days(15), by = "day"))
    
  }) |> rbindlist()

for (t in times$V1) {
  
  message(t)
  if (t == times$V1[1]) {
    
    print(1)
    temp <- ReadNetCDF("/scratch/w40/pc2687/2t_daily_deseasoned.nc", vars = "t2m", subset = list(time = as_date(t)))
    
    temp[, time := NULL]
    
  } else {
    
    field <- rbind(temp, 
          ReadNetCDF("/scratch/w40/pc2687/2t_daily_deseasoned.nc", vars = "t2m", subset = list(time = as_date(t))) |> 
            _[, time := NULL])  |> 
      _[, .(t2m = sum(t2m)), by = .(longitude, latitude)]
  }
}

field <- field[, t2m := t2m/nrow(times)]
               
readr::write_rds(field, "data/2t_7m.rds")
field |> 
  ggplot(aes(longitude, latitude)) +
  # scattermore::geom_scattermore(aes(color = t2m)) +
  geom_contour_fill(aes(z = t2m, fill = after_stat(level))) +
  scale_fill_divergent_discretised(guide = guide_colorbar(barwidth = 20,
                                               barheight = 0.5)) +
  geom_sf(data = map, inherit.aes = FALSE, 
          fill = NA, color = "grey", linewidth = 0.2) +
  coord_sf(expand = FALSE) +
  labs(x = NULL, y = NULL, fill = NULL,
       subtitle = "Anomaly of t2m using 30 days centered in day 180 after eruption") +
  theme_minimal() +
  theme(legend.position = "bottom")

```



Reads the year previous to the eruption and calculate the mean base temperature for all eruptions.

```{r}
times_base <- purrr::map(strong$ini_date, function(d) {
    
    # c(d - days(1), d + days(1))
    times <- list(seq.Date(d - days(365), d - days(1), by = "days"))
    
  }) |> rbindlist() |> 
  _[, .(time = unique(V1))]

for (t in times_base$time) {
  
  message(as_date(t))
  if (t == times_base$time[1]) {
    
    print(1)
    temp <- ReadNetCDF("/scratch/w40/pc2687/2t_daily_deseasoned.nc", vars = "t2m", subset = list(time = as_date(t))) |> 
      _[, time := NULL]
    
  } else {
    
    field_base <- rbind(temp, 
          ReadNetCDF("/scratch/w40/pc2687/2t_daily_deseasoned.nc", vars = "t2m", subset = list(time = as_date(t))) |> 
            _[, time := NULL])  |> 
      _[, .(t2m = sum(t2m)), by = .(longitude, latitude)]
  }
}

field_base <- field_base[, t2m := t2m/nrow(times_base)]

readr::write_rds(field_base, "data/2t_base_1y.rds")
               
field_base |> 
  ggplot(aes(longitude, latitude)) +
  # scattermore::geom_scattermore(aes(color = field$t2m - t2m)) +
  geom_contour_fill(aes(z = field$t2m - t2m, fill = after_stat(level)),
                    breaks = seq(-0.2, 0.2, 0.02)) +
  scale_fill_divergent_discretised(guide = guide_colorbar(barwidth = 20,
                                               barheight = 0.5)) +
  geom_sf(data = map, inherit.aes = FALSE, 
          fill = NA, color = "grey", linewidth = 0.2) +
  coord_sf(expand = FALSE) +
  labs(x = NULL, y = NULL, fill = NULL,
       subtitle = "Anomaly of t2m using 30 days centered in day 180 after eruption") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
```{r}

field_base[field, on = c("latitude", "longitude")] |> 
  _[latitude %between% c(-30, 30), diff := i.t2m - t2m] |> 
    ggplot(aes(longitude, latitude)) +
  # scattermore::geom_scattermore(aes(color = field$t2m - t2m)) +
  geom_contour_fill(aes(z = diff, fill = after_stat(level))) +
  scale_fill_divergent_discretised(guide = guide_colorbar(barwidth = 20,
                                               barheight = 0.5)) +
  geom_sf(data = map, inherit.aes = FALSE, 
          fill = NA, color = "grey", linewidth = 0.2) +
  coord_sf(expand = FALSE, ylim = c(-30, 30)) +
  labs(x = NULL, y = NULL, fill = NULL,
       subtitle = "Anomaly of t2m using 30 days centered in day 210 after eruption") +
  theme_minimal() +
  theme(legend.position = "bottom")
```



## Number of pixels over/below percentiles

```{r}
files <- Sys.glob("data/2t_gt_p*")

n_px <- purrr::map(files, function(f) {
  
  meta <- unglue(f, "data/2t_gt_{percentile}_npx.rds")  
  
  readr::read_rds(f) |>
    setDT() |>
    _[, let(time = floor_date(time, "day"),
            percentile = meta[[1]][["percentile"]])]
  
  
}) |> 
  rbindlist() |> 
  _[, n_px := fifelse(percentile %in% c("p1", "p5"), 1038240 - n_px, n_px)]

```


```{r}
get_series_np <- function(ini_date, before = 1, after = 3, p = "p1") {
  
  start_time <- ini_date - years(before)
  end_time <- ini_date + years(after)
  temp <- n_px[time %between% c(start_time, end_time) & percentile == p]
  
  list(time = temp$time,
       n_px = temp$n_px)
  
}

calculate_base <- function(temperature_px, ini_date, period = 12, p = "p1") {
  
  
  start_period <- lubridate::as_datetime(ini_date) - months(period)
  
  as.numeric(temperature_px[time %between% c(start_period, ini_date) & percentile == p,
                            .(base_px = mean(n_px))])
  
}
```


```{r}
t2_p1 <- strong |>
  _[, let(base = calculate_base(n_px, ini_date, p = "p1")),
    by = .(volcano, ini_date)] |>
  _[,  get_series_np(ini_date, after = 2, p = "p1"),
    by = .(volcano, ini_date)] |>
  
  strong[i = _, on = c("volcano", "ini_date")] |>
  # _[]
  _[, let(diff = n_px - base, 
          id = paste0(volcano, "_", ini_date),
          lag =  as.numeric(difftime(time, ini_date))/2592000)]


t2_p1 |>
  ggplot(aes(lag, diff)) +
  geom_vline(xintercept = 0, color = "darkorange") +
  geom_hline(yintercept = 0,color = "grey") +
  geom_line(aes(group = id)) +
  geom_point(data = t2_p1[, .SD[diff == max(diff)], by = .(id)], color = "red") +
  geom_smooth(alpha = 0.5, linewidth = 0.5) +
  # geom_text(data = new_strong, x = 0, y = -0.4, hjust = 0, size = 3,
  #           aes(label = paste0("SO2: ", so2_kt, "\nAlt: ", p_alt_obs, "\nVEI: ", vei))) +
  facet_wrap(~id, ncol = 3) +
  labs(x = "Months since eruption",
       y = "# pixels with 2t below 1th percentile") +
  theme_minimal()

```

```{r}
t2_p99 <- strong |>
  _[, let(base = calculate_base(n_px, ini_date, p = "p99")),
    by = .(volcano, ini_date)] |>
  _[,  get_series_np(ini_date, after = 2, p = "p99"),
    by = .(volcano, ini_date)] |>
  
  strong[i = _, on = c("volcano", "ini_date")] |>
  # _[]
  _[, let(diff = n_px - base, 
          id = paste0(volcano, "_", ini_date),
          lag =  as.numeric(difftime(time, ini_date))/2592000)]


t2_p99 |>
  ggplot(aes(lag, diff)) +
  geom_vline(xintercept = 0, color = "darkorange") +
  geom_hline(yintercept = 0,color = "grey") +
  geom_line(aes(group = id)) +
  geom_point(data = t2_p99[, .SD[diff == max(diff)], by = .(id)], color = "red") +
  # geom_smooth(method = "lm", alpha = 0.5, linewidth = 0.5) +
  # geom_text(data = new_strong, x = 0, y = -0.4, hjust = 0, size = 3,
  #           aes(label = paste0("SO2: ", so2_kt, "\nAlt: ", p_alt_obs, "\nVEI: ", vei))) +
  facet_wrap(~id, ncol = 3) +
  labs(x = "Months since eruption",
       y = "# pixels with 2t over 99th percentile") +
  theme_minimal()
```

```{r}
t2_p1[, .SD[diff == max(diff)], by = .(id)] |> 
  _[, time]

ReadNetCDF("/scratch/w40/pc2687/gt_p95/2t/1p/daily_2t_gt_p1_199212.nc", vars = "t2m", 
           subset = list(time = ymd("1992-12-21"))) |> 
  _[, t2m := fifelse(t2m == 0, 1, 0)] |> 
  _[t2m == 1] |> 
  ggplot(aes(longitude, latitude)) +
  geom_raster(fill = "cyan4") +
  geom_sf(data = map, inherit.aes = FALSE, fill = NA, color = "grey", linewidth = 0.2) +
  labs(title = "Points with T2 below 1th percentile",
       subtitle = "1992-12-21") +
  theme_minimal()
```

```{r}
t2_p99[, .SD[diff == max(diff)], by = .(id)] |> 
  _[, time]

ReadNetCDF("/scratch/w40/pc2687/gt_p95/2t/99p/daily_2t_gt_p99_200911.nc", vars = "t2m", 
           subset = list(time = ymd("2009-11-12"))) |> 
  # _[, t2m := fifelse(t2m == 0, 1, 0)] |> 
  _[t2m == 1] |> 
  ggplot(aes(longitude, latitude)) +
  geom_raster(fill = "cyan4") +
  geom_sf(data = map, inherit.aes = FALSE, fill = NA, color = "grey", linewidth = 0.2) +
  labs(title = "Points with T2 over 99th percentile",
       subtitle = "2009-11-12") +
  theme_minimal()
```

